### 2023-09-06 article-pl.tex
Initial version of article

% !TEX encoding = UTF-8 Unicode

%%%%%%%%%%%%%%%%%%%% author.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample root file for your "contribution" to a proceedings volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[conference]{IEEEtran}
%
% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

% to typeset URLs, URIs, and DOIs
% https://www.overleaf.com/learn/latex/Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    pdftitle={Overleaf Example},
    pdfpagemode=FullScreen,
    }

\def\UrlFont{\rmfamily}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[textsize=scriptsize]{todonotes}
\usepackage{gensymb}
\usepackage{array,multirow}
\usepackage{caption} 
\captionsetup[table]{skip=10pt}
\usepackage{makecell}

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{bm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\usepackage{graphicx}
\graphicspath{ {./img/} }

\renewcommand{\labelenumii}{\theenumii}
\renewcommand{\theenumii}{\theenumi.\arabic{enumii}.}

\begin{document}
%
\title{Duckietown linie folower PID and AI}

\author{
\IEEEauthorblockN{1\textsuperscript{st} Marek Długosz}
\IEEEauthorblockA{\textit{AGH University of Science and Technology}\\
Cracow, Poland \\
mdlugosz@agh.edu.pl \\
0000-0001-6827-9149}\\[0.4cm]
\IEEEauthorblockN{3\textsuperscript{rd} Marcin Szelest}
\IEEEauthorblockA{\textit{IEEE Senior Member}}
\IEEEauthorblockA{\textit{AGH University of Science and Technology} \\
Cracow, Poland \\
mszelest@agh.edu.pl\\
0000-0002-0522-1270}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Paweł Skruch}
\IEEEauthorblockA{\textit{IEEE Senior Member}}
\IEEEauthorblockA{\textit{AGH University of Science and Technology} \\
Cracow, Poland \\
pawel.skruch@agh.edu.pl\\
0000-0002-8290-8375}\\
\IEEEauthorblockN{4\textsuperscript{th} Artur Morys-Magiera}
\IEEEauthorblockA{\textit{AGH University of Science and Technology} \\
Cracow, Poland \\
amorys@student.agh.edu.pl\\
0000-0002-2137-8841}
}

\maketitle              % typeset the title of the contribution

\begin{abstract}

\end{abstract}

% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\begin{IEEEkeywords}
\end{IEEEkeywords}
%
\section{Introduction}

\section{Conclusion}\label{sec:conclusion}

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}

### 2023-09-12 article-pl.tex
Polish version of article

@@
 \usepackage{caption} 
+\usepackage{subcaption}
 \captionsetup[table]{skip=10pt}
@@
 %
-\title{Duckietown linie folower PID and AI}
+\title{Zastosowanie regulatora PID i CNN do sterowania robotem Duckiebot}
+% \title{Application of PID and CNN controllers to control the Duckiebot robot}
 
@@
 \begin{IEEEkeywords}
+% autonomouse robot, PID controller, convolutional neural network, line follower
+autonomiczny robot, regulator PID, konwolucyjne sieci neuronowe, line follower
 \end{IEEEkeywords}
@@
 \section{Introduction}
+Ostatnimi latami można zaobserwować swoisty renesans w wykorzystaniu sztucznych sieci neuronowych.
+Powstały i są ogólnie dostępne różnego rodzaju architektury sieci, udoskonalono ich metody trenowania.
+Wszystko to sprawiło, że  sztuczne sieci neuronowe są coraz częściej, z sukcesem, wykorzystywane w układach automatycznego sterowania \cite{algor2030973, 786109}.
+Zastosowanie sieci neuronowych w układach automatyki nie jest ich typowym wykorzystaniem ponieważ bardzo często sterowanie może być realizowane z wykorzystaniem dobrze znanych i opisanych klasycznych algorytmów sterowania (PID, LQR, MPC itp.).
+Daje to okazję do porównania działa dwóch różnych technologii, znalezienia ich wad oraz plusów.
+W procesie kształcenia studentów projekt i wykonanie przez nich układu sterowania z wykorzystaniem klasycznych algorytmów sterowania oraz sztucznych sieci neuronowych jest bardzo interesujące zarówno z punktu naukowego jak i dydaktycznego.
 
-\section{Conclusion}\label{sec:conclusion}
+Istotnym elementem kształcenia studentów podnoszącym atrakcyjność samych zajęć, dającym możliwość rozwiązywania problemów praktycznych jest wykorzystanie rzeczywistych robotów.
+Jednym z takich projektów dostarczających kompletną platformę robotyczną (sprzęt, oprogramowanie, symulator, materiały dydaktyczne) jest projekt Duckietown.
+Projekt Duckietown powstał w 2016 na uniwersytecie Massachusetts Institute of Technology (MIT) w ramach zajęć dla studentów \cite{paull2017duckietown}.
+Jest to otwarty projekt, którego celem jest promowanie edukacji i badań związanych z autonomią robotów, zwłaszcza w kontekście samochodów autonomicznych.
+W ramach projektu, powstały roboty Duckiebot, makiety miast Duckietown po których poruszają się roboty, oraz dodatkowa infrastrukturę, taką jak np.
+Watchtower, wspomagającą proces nawigacji.
+Projekt obejmuje także stworzenie oprogramowania symulacyjnego o nazwie Duckiebot-GYM, które umożliwia symulację pracy robotów w wirtualnym środowisku \cite{gym_duckietown}.
+Obecnie dostępna jest już druga generacja robotów Duckiebot, w której do sterowania wykorzystuje się komputer Jetson Nano.
+Wykorzystanie platformy Jetson Nano otwiera drzwi do wykorzystania zaawansowanych sieci neuronowych z dedykowanym wsparciem GPU do sterowania robotem Duckiebot.
+
+Artykuł zorganizowany jest następująco, w sekcji \ref{sec:problem-definition} zdefiniowano problem do rozwiązania, w sekcji \ref{sec:student-activity} opisano aktywności studentów realizowane w trakcie rozwiązywania  problemu.
+Sekcja \ref{sec:robot-duckiebot} zawiera krótki projektu Duckietown oraz konstrukcji i sposobu sterowania robotem Duckiebot, w sekcjach \ref{sec:pid-controller} i \ref{sec:nn-controller} opisano sposób rozwiązania problemu z wykorzystaniem regulatora PID i regulatora wykorzystującego sztuczne sieci neuronowe.
+W sekcji \ref{sec:practical-info} podano kilka praktycznych wskazówek pomocnych w trakcie realizacji zadania i w ostaniej sekcji \ref{sec:conclusion} zawarto wnioski.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Definicja problemu}\label{sec:problem-definition}
+Ramach procesu dydaktycznego, studenci mają za cel zaprojektować układ automatycznego sterowania robotem o napędzie różnicowym który realizuje funkcjonalność poruszania się po wyznaczonej trasie.
+Trasa robota oznaczona jest linią o kolorze (np.
+kolor czerwony, żółty itp.) różnym od podłoża (najczęściej w kolorze czarnym) po którym porusza się robot.
+Kształt trasy stanowi zamknięty kontur.
+Układ sterowania powinien generować sygnał sterowania aby minimalizować całkowity błąd odchyłki robota od zadanej trasy.
+Tego typu zadanie, jest typowym zdaniem implementowanym w różnego rodzaju robotach autonomicznych i często jest określane jako \emph{line follower}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Studenckie aktywności}\label{sec:student-activity}
+W ramach zdefiniowanego problemu w sekcji \ref{sec:problem-definition} studenci będą musieli się wykazać odpowiednim przygotowaniem teoretycznym oraz umiejętnościami praktycznymi takimi jak: 
+znajomość zasady działania regulatora PID, znajomość zasady działania sztucznych sieci neuronowych, znajomość operacji związanych z przekształceniami obrazów, umiejętność implementacji regulatora PID w wersji dyskretnej, umiejętność implementacji wybranych struktur sieci neuronowych, umiejętność implementowania operacji przekształceń obrazów, znajomość zagadnień związanych z doborem parametrów regulatora PID, znajomość zagadnień związanych z uczeniem sieci neuronowych i przygotowaniem danych do ich uczenia, ocena poprawności procesu uczenia sieci.
+Z pozoru prosty algorytm sterowania nieskomplikowanym robotem realizujący funkcjonalność \emph{line follower} generuje szereg zadań których wykonanie wymaga odpowiedniego przygotowania teoretycznego jak i praktycznego studentów.
+Należy również zauważyć, że problemy które studenci muszą rozwiązać w trakcie realizacji projektu są typowymi problemami które występują w przy programowaniu innego typu robotów autonomicznych.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Robot Duckiebot}\label{sec:robot-duckiebot}
+Robot Duckiebot jest niewielkim, dwu kołowym robotem zdolnym do poruszania się po płaskich nawierzchniach.
+Do napędu robota wykorzystano dwa moduły zawierające: silnik elektryczny, przekładnię i enkoder.
+Robot Duckiebot wyposażony jest w następujące sensory: kamera RGB, czujnik odległości oraz czujnik bezwładnościowy IMU \cite{gupta2022low}.
+Do sterowanie robotem wykorzystany jest komputer JetsonNANO.
+Dodatkowo robot Duckiebot wyposażony jest w cztery diody RGB które mogą być wykorzystany np.
+do sygnalizowania stanu w jakim aktualnie znajduje się robot.
+Robot Duckiebot jest zaprojektowany do autonomicznego sterowania i może poruszać się bez udziału człowieka, na podstawie danych z kamery i odpowiednich algorytmów sterowania.
+Na rysunku\;\ref{fig:duckiebot-3d} przedstawiono robota Duckiebot w trzech rzutach.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.9\columnwidth]{duckiebot-blue-3d}
+    \caption{Robot Duckiebot (\url{http://duckietown.org}).}
+    \label{fig:duckiebot-3d}
+\end{figure}
+
+Konsekwencją zastosowania dwóch niezależnych silników elektryczny w konstrukcji robota Duckiebot jest sposób w jaki kontroluje się kierunek poruszania się robota.
+Robot Duckiebot wykonuje zakręty zmieniając prędkości obrotowe silników elektrycznych w zależności od tego w jaką stronę chce zakręcić oraz jak duży ma być promień skrętu.
+Do sterowania robotem wykorzystywane są dwa sygnały: $v$-wartość prędkości liniowej oraz $\omega$-wartość prędkości obrotowej wzdłuż osi pionowej Z robota, zob. rys. \ref{fig:duckiebot-kinematics}.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.8\columnwidth]{dt-kinematic}
+    \caption{Układ kinematyczny robota Duckiebot.}
+    \label{fig:duckiebot-kinematics}
+\end{figure}
+
+Znając wartości $(v, \omega)$ (wyznaczone np. przez regulator) możemy obliczyć prędkości obrotowe kół robota z wzorów \eqref{eq:omega-L} i \eqref{eq:omega-R}:
+
+\begin{eqnarray}
+	\omega_L & = & \cfrac{v-L\omega}{R} \label{eq:omega-L} \\
+	\omega_R & = & \cfrac{v+L\omega}{R} \label{eq:omega-R}
+\end{eqnarray}
+\\
+Tego typu układ sterowania robotem Duckiebot jest bardzo często wykorzystywany w robotyce i nosi nazwę \emph{differential drive robot} \cite{siegwart2011introduction}.
+Zaletą takiego układu kinematycznego ruch robota Duckiebot jest prostota konstrukcji (brak mechanizmów osi skrętnych kół) natomiast wadą jest konieczność ciągłego kontrolowania prędkości obrotowej kół robota ($\omega_R$, $\omega_L$) tak aby poruszał się on po zadanej trasie.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Klasyczna realizacja z regulatorem PID}\label{sec:pid-controller}
+Klasyczny układu sterowania robotem Duckiebot wykorzystuje regulator PID.
+Zadniem układu sterowania jest generowanie takiego sygnału sterowania $u_i$ aby minimalizować wartość sygnału błędu $e_i$, gdzie $i$-kolejne chwile czasowe.
+Na rysunku \ref{fig:pid-pipeline} przedstawiona schemat blokowy układu sterowania robotem Duckiebot z wykorzystaniem regulatora PID oraz kamery robota jako sensorem.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.8\columnwidth]{PipeLinePID}
+    \caption{Schemat układu sterowania robotem Duckiebot z wykorzystaniem regulatora PID.}
+    \label{fig:pid-pipeline}
+\end{figure}
+
+Sygnał błędu $e_i$ zdefiniowany jest jako różnica pomiędzy środkiem osi pionowej obrazu z kamery a punktem znajdującym się w analizowanym obszarze obrazu zawierającym linię wzdłuż której powinien poruszać się robot Duckiebot, zob.
+rys.\;\ref{fig:error-definition}.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.95\columnwidth]{ErrorDefinition.png}
+    \caption{Określenie błędu dla układu sterowania realizującego funkcjonalność \emph{line follower} dla robota Duckiebot.}
+    \label{fig:error-definition}
+\end{figure}
+
+Układ sterowania robota Duckiebot jest systemem dyskretnym (zaimplementowanym na komputerze JetsoNANO) dlatego należy wykorzystać wersję dyskretną regulatora PID.
+Uwzględniając rodzaj napędu robot Duckiebot (napęd różnicowy \emph{differential drive robot}) sygnał sterowania $u_i$ wygenerowany przez regulatora PID odpowiada prędkości obrotowej $\omega_i$ wzdłuż osi Z robota.
+
+\subsection{Regulator PID}
+Jednym z najlepiej poznanych i najczęściej stosowanych regulatorów w układach automatycznego sterowania jest regulator PID.
+Wynika to przede wszystkim z jego uniwersalności, łatwości realizacji zarówno w formie analogowej jak i cyfrowej oraz w miarę prostego i intuicyjnego sposobu strojenia parametrów regulatora PID.
+Regulator PID w wersji dyskretnej który zastosowano w układzie sterowania robotem Duckietown dany jest równaniem:
+
+\begin{equation}
+u_i = K_p e_i + K_i \sum_{i=0}^k{e_i \Delta t} + K_d \cfrac{e_i-e_{i-1}}{\Delta t}
+\label{eq:pid-dis}
+\end{equation}
+%https://github.com/duckietown/mooc-exercises/blob/c476a67fea9836beab99eed23cfc1a83c77af7bb/modcon/solution/05-PID-Control/PID_controller.ipynb
+gdzie: $K_p$ wzmocnienie części proporcjonalnej, $K_i$ wzmocnienie części całkującej, $K_d$ wzmocnienie części różniczkującej, $e_i$ błąd w $i$-tej chwili czasu, $\Delta t$ krok czasowy \cite{aastrom2021feedback}.
+
+\subsection{Pomiar błędu}
+Wartość sygnału błędu $e_i$ wyznaczana jest na podstawie obrazów przesyłanych przez kamerę robota Duckiebot.
+W celu określenia wartości sygnału $e_i$ konieczne jest wykonanie kilku operacji na przesłanych obrazach (zob.
+rys.
+\ref{fig:pid-pipeline}).
+Pierwszym operacją jest konwersja obrazu z przestrzeni barw RGB to przestrzeni HSV (\emph{Hue}, \emph{Saturation}, \emph{Value}).
+Takie przekształcenie przestrzeni barw obrazów ma na celu przede wszystkim lepsze rozróżnianie i rozpoznawanie kolorów.
+Kolejną operacją jest znalezienie na obrazie lini o określonym kolorze wzdłuż której ma poruszać się robot Duckiebot.
+Taka operacja ma na celu pozostawienia na obrazie tylko lini wyznaczającej trasę robotą i usunięcie z obrazu wszystkich pozostałych elementów (np.
+tło).
+Kolejna operacją jest ograniczenie obszaru analizowanego poprzez wykonanie operacji przycięcia. Operacja ta ma na celu dokładne znalezienie środka lini wzdłuż której ma poruszać się robot Duckiebot. 
+Środek lini znajdywany jest jako współrzędne środka ciężkości otrzymanego, w wyniku poprzednich przekształceń, lini wyznaczającej trasę robota Duckiebot.
+Na rysunku \ref{fig:img-transformation-summary} pokazano rezultat poszczególnych przekształceń wykonywanych na obraza z kamery robota w celu wyznaczenia wartości błędu położenia $e_i$.
+
+\begin{figure}[hbt!]
+    \centering
+    \includegraphics[width=.95\columnwidth]{img-transformation.png}
+    \caption{Operacje wykonywane na obrazach z kamery robota w celu wyznaczenie błędu położenia $e_i$.}
+    \label{fig:img-transformation-summary}
+\end{figure}
+
+Na końcu wartości błędu położenia $e_i$ jest normalizowana do przedziału $e_i \in [-1;\;1]$. Ma to na celu uniezależnieniu się od rozmiarów (szerokość, wysokość) obrazów przesyłanych z kamery.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Realizacja z wykorzystaniem sieci neuronowych}\label{sec:nn-controller}
+Drugi ze sposobów generowania sygnału sterowania $u_i$ wykorzystuje konwolucyjną sieć neurnową (\emph{Convolutional Neural Network} CNN).
+Schemat blokowy układu regulacji wykorzystującego sieci neuronowe przedstawiono na rys.
+\ref{fig:cnn-pipeline}.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.8\columnwidth]{NNPipeline3}
+    \caption{Schemat układu sterowania robotem Duckiebot z wykorzystaniem konwolucyjnych sieci neurnowych.}
+    \label{fig:cnn-pipeline}
+\end{figure}
+
+Konwolucyjne sieci neuronowe są bardzo skuteczne szczególnie w zadaniach rozpoznawania wzorców i przetwarzania obrazów.
+Sieci CNN wykorzystują wielokrotnie operację konwolucji której użycie ma na celu wydobywanie z obrazów wejściowych charakterystycznych cech (krawędzie, tekstury, kształty) na podstawie których klasyczna sieć neuronowa może przeprowadzić wnioskowanie "co znajduje się na obrazie".
+Konwolucyjne sieci neuronowe są  stosowane w różnych dziedzinach, takich jak rozpoznawanie obrazów, analiza tekstu, przetwarzanie mowy, rozpoznawaniem twarzy, samochodów, diagnozowaniem chorób na podstawie obrazów medycznych, analizą obrazów satelitarnych itp.
+\cite{li2021survey}, \cite{rawat2017deep}, \cite{almasi2020robust}.
+
+W przypadku wykorzystania do sterowania robotem Duckiebot sieci CNN, sieć taka na wejściu otrzymuje obrazy z kamery (po wykonaniu kilku przekształceń) na podstawie których generuje sygnał sterowania $u_i$ który jest interpretowany jako  $\omega_i$ prędkość obrotowa wzdłuż osi Z robota, zob.
+rys.
+\ref{fig:cnn-pipeline}.
+
+\subsection{Sieć neuronowa}
+Proponowana struktura konwolucyjnej sieci neuronowej przedstawiona jest na rys. \ref{fig:cnn-structure} i należy zaznaczyć iż jest to jedna z możliwych do wykorzystania struktur sieci.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.5\columnwidth]{cnn2}
+    \caption{Struktura konwolucyjnej sieci neuronowej wykorzystanej do sterowania robotem Duckiebot.}
+    \label{fig:cnn-structure}
+\end{figure}
+
+Sieć składa się z pierwszej warstwy konwolucyjnej, następną warstwa jest warstwa \emph{MaxPooling}.
+Jej zdaniem jest zredukowanie rozmiaru danych wejściowych.
+Kolejne  warstwy sieci to warstwa konwolucyjnej i warstwa MaxPooling - taki układ warstw jest powtórzony dwukrotnie.
+Za każdym razem warstwy konwolucyjne ekstrahują kolejne cechy obrazu wejściowego istotnie z punktu wiedzenia sterowania robotem Duckiebot, natomiast warstwy \emph{MaxPooling} redukują rozmiar danych.
+Zadaniem następnej warstwy \emph{Dropout} losowe wyłączanie poszczególnych neuronów sieci, tak aby zapobiec zjawisku \emph{overfitting} oraz wspomóc zdolność generalizacji sieci neuronowej.
+Przed przesłaniem danych na wejście właściwej sieci neuronowej muszą one zostać spłaszczone co jest zadaniem warstwy \emph{Flatten}.
+Przekształca ona dane wielowymiarowe w jednowymiarowy wektor wejściowy do następnej warstwy \emph{Dense}.
+Warstwa \emph{Dense} jest jest odpowiedzialna za wykonywanie klasyfikacji, regresji lub innych zadań uczenia maszynowego.
+Wyjściem z sieci CNN jest wartość sygnału $\omega$ z przedziału $[-8\: ; \:8]$ dlatego funkcją aktywacji jest z wyjścia warstwy \emph{Dense} jest funkcja \emph{tanh}.
+
+\subsection{Przygotowanie danych}
+Każda sieć neuronowa jest tak dobra jak dobrymi danymi została wytrenowana.
+Stąd też bardzo ważnym etap pracy z sieciami neuronowymi jest przygotowanie dobrego zbioru danych uczących.
+W przypadku robota Duckiebot i zadania które ma realizować (\emph{line follower}) dane uczące dla sieci składają się z par: obraz z kamery i odpowiadająca mu prawidłowa wartość $\omega$, zob.
+rys.
+\ref{fig:cnn-pipeline}.
+Dane konieczne do trenowania sieci można uzyskać sterując robotem Duckiebot ręcznie (np.
+za pomocą gamepad) lub wykorzystać opracowany wcześniej układ sterowania z regulatorem PID dodając skrypt zapisujący dane.
+
+Oprócz odpowiedniej ilości zalogowany danych do uczenia sieci neuronowej, ważne jest aby dane te były odpowiednio różnorodne - czyli aby obejmowały jak największą liczbę przypadków, sytuacji, zdarzeń w jakich może znaleźć się robot Duckiebot.
+Na rysunku \ref{fig:cnn-data-hist1} przedstawiono histogram przykładowego zbioru danych jaki został zalogowany.
+
+\begin{figure}[h]
+	\begin{subfigure}{.475\columnwidth}
+    	\centering
+    	\includegraphics[width=1\columnwidth]{h1}
+    	\caption{}
+    	\label{fig:cnn-data-hist1}
+    \end{subfigure}
+    \begin{subfigure}{.475\columnwidth}
+    	\centering
+    	\includegraphics[width=1\columnwidth]{h2}
+    	\caption{}
+    	\label{fig:cnn-data-hist2}
+    \end{subfigure}
+
+    \caption{Histogram zalogowanych danych przed analizą a) i po analizie b).}
+    \label{fig:cnn-data-hist}
+\end{figure}
+
+Wyraźnie widać, że dominującą wartością w tym zbiorze danych dla zmiennej $\omega$ są wartości w bliskie wartości $0$ i jeśli użyć takich danych do trenowania sieci CNN która ma sterować robotem Duckiebot to prawdopodobnie robot bardzo dobrze jeździł by prosto, natomiast miał by problemy z zakrętami.
+W takim przypadku, po analizie danych należy usunąć ze zbioru uczącego dane których wartości znacząco dominują nad innymi wartościami otrzymując bardziej reprezentatywny zbiór danych uczących, zob.
+rys.
+\ref{fig:cnn-data-hist2}.
+
+Podobnie jak w przypadku sterowania robotem Duckiebot z wykorzystaniem regulatora PID tak i w przypadku sterowania robotem Duckiebot z wykorzystaniem sieci neuronowych konieczne jest odpowiednie przygotowania obrazów z kamery.
+Poszczególne etapy przetwarzania obrazów z kamer robota Duckiebot przedstawiono na rysunku \ref{fig:cnn-pipeline} (bloki o kolorze żółtym) i są to: przycinanie obrazu do interesującego nas obszaru, konwersja do przestrzeni barw YUV (Y-luminacja, UV-barwa kolorów, chrominancja), filtrowanie z wykorzystaniem filtru Gauus-a, zmiana rozmiaru obrazu i normalizacja wartości poszczególnych kanałów obraz do przedziału $[0\;1]$.
+Rezultat poszczególnych operacji można zobaczyć na rysunku \ref{fig:cnn-image-prepare}.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.95\columnwidth]{nn_prepare_image}
+    \caption{Operacje wykonywane na obrazach z kamery robota Duckiebot sterowanym przed zastosowanie regulatora CNN.}
+    \label{fig:cnn-image-prepare}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Praktyczne wskazówki}\label{sec:practical-info}
+Oprogramowanie robotów Duckietown zostało zaimplementowane z wykorzystaniem framework-a Robot Opertion System (\emph{ROS}) \cite{quigley2015programming}.
+Jest on obecnie jednym z najpopularniejszym framework-ów do programowania rożnego rodzaju robotów.
+Konsekwencją wybrania framework-a ROS jest wybór języka programowania i w przypadku robotów Duckiebot jest to Python.
+Projekt Duckietown definiuje również sposób tworzenia własnego oprogramowania dla robotów i wymusza stosowanie technologi konteneryzacji.
+Jest to bardzo dobra praktyka zgodna z ówczesnymi trendami tworzenia oprogramowania.
+Dokładny opis w raz z przykładami tworzenia oprogramowania dla robotów Duckie jest bardzo dobrze opisany w dokumentacji projektu \cite{Daniele:ff}.
+Wykorzystanie framework-a ROS umożliwia również efektywnie tworzenie i testowania oprogramowania bez konieczności ciągłego wygrywania kolejnych wersji do robota Duckiebot.
+Konfigurując odpowiednio komputer host-a, tak aby łączy się z serwerem ROS-na uruchomionym na wybranym robocie Duckiebot można w sposób wygodny i efektywny implementować i testować własne oprogramowanie, zob.
+rys.
+\ref{fig:env-develop}.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=.8\columnwidth]{controll_schema2}
+    \caption{Konfiguracja środowiska deweloperskiego dla robota Duckiebot.}
+    \label{fig:env-develop}
+\end{figure}
+
+Proponowane regulatory należy zaimplementować jako węzły ROS-a.
+Węzły powinny odczytywać obrazy z kamery robota (przesyłane jako wiadomości ROS) a jako wyjście generować sygnał sterowania $\omega$.
+Dla uproszczenia sterowania robotem Duckiebot można przyjąć, że porusza się on ze stała prędkością liniową $v=const$.
+
+Konwolucyjne sieci neuronowe można tworzyć, trenować oraz testować wykorzystując moduły \emph{keras} i \emph{tensorflow} języka Python.
+Zaletą modułów \emph{keras} i \emph{tensorflow} jest również bardzo dobra dokumentacja z bardzo duża ilością przykładów.
+
+Należy zwrócić również uwagę prawidłowe przygotowanie danych do uczenia sieci neuronowych, każdemu obrazowi z kamery musi odpowiadać jedna, poprawna wartość sygnału $\omega$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Wnioski}\label{sec:conclusion}
+Regulator PID jest stosowany z powodzeniem w wielu układach sterowania a jego implementacja jest stosunkowo prosta.
+Dla tego typu regulatorów istnieje również szereg metod i algorytmów doboru parametrów regulatora.
+Regulatory PID nie są wolne też od wad.
+Jedną z nich jest znajomość, choćby przybliżonego, modelu procesu jakim chcemy sterować.
+Zachodzi więc konieczność identyfikacji zarówno struktury modelu procesu jak i jego parametrów.
+Zadania identyfikacji nie są zadaniami prostymi, wymagają dużej wiedzy odnośnie istoty samego procesu.
+Istnieją również metody identyfikacji modelów procesów bazujące na wynikach eksperymentów praktycznych, ale w tym przypadku może się okazać, że przeprowadzanie takiego eksperymentu nie zawsze jest możliwie.
+Stosują regulator PID należy również mieć świadomość, że został on opracowany dla procesów których działanie można opisać modelami liniowymi.
+Niestety, działanie zdecydowanej większości systemów dynamicznych jest opisane modelami nieliniowymi.
+Konsekwencją tego jest fakt, że w takim przypadku regulator PID działa wykorzystując liniowe przybliżenia systemów nieliniowych - co może być przyczyną różnych błędów, niedokładności itp.
+
+W przeciwieństwie do klasycznego regulatora PID, regulatory wykorzystujące sztuczne sieci neuronowe nie potrzebują znajomości modelu matematycznego procesu którym sterują oraz jego parametrów.
+Możliwość projektowania różnych architektur sieci neuronowych, takich jak konwolucyjne, rekurencyjne czy głębokie sieci neuronowe, umożliwia dostosowanie regulatora neuronowego do konkretnego procesu którym mają sterować.
+Z drugiej strony, mnogość architektur sieci neuronowych, ich budowa powoduje, że nigdy nie mamy pewności czy dana struktura sieci neurnowej jest optymalna.
+Dobór parametrów regulatorów neuronowych odbywa się w sposób automatyczny poprzez wykorzystanie odpowiednich algorytmów uczenia sieci.
+Kluczowym elementem mającym wpływ na poprawność pracy regulatora neuronowego są dane jakim jest uczona sieć neuronowa.
+Wadą regulatorów wykorzystujących sieci neuronowe jest brak możliwości wykazania stabilności działania układów którymi sterują.
+W przypadku regulatora PID, pomimo wykorzystywania przybliżonych modeli procesu bardzo często można udowodnić, że zamknięty układ sterowania będzie pracował stabilnie w każdych lub w pewnym zakresie wartości zmiennych.
+Niestety takiej analizy nie da się przeprowadzić w przypadku regulatorów neuronowych.
+
+Podsumowując implementacja dwóch różnych regulatorów do realizacji tego samego zadnia daje możliwość poznania zalet i wad każdego z nich.
+Studenci w trakcie rozwiązywania postawionego problemu muszą wykazać się znajmością teoretyczną  zagadnień oraz, co jest szczególnie istotne w przypadku inżynierów, nabywają umiejętności praktycznego stosowania wiedzy teoretycznej.
 

### 2023-09-13 article-en.tex
English version of article

% !TEX encoding = UTF-8 Unicode

%%%%%%%%%%%%%%%%%%%% author.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample root file for your "contribution" to a proceedings volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[conference]{IEEEtran}
%
% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%

% to typeset URLs, URIs, and DOIs
% https://www.overleaf.com/learn/latex/Hyperlinks
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    pdftitle={Overleaf Example},
    pdfpagemode=FullScreen,
    }

\def\UrlFont{\rmfamily}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[textsize=scriptsize]{todonotes}
\usepackage{gensymb}
\usepackage{array,multirow}
\usepackage{caption} 
\usepackage{subcaption}
\captionsetup[table]{skip=10pt}
\usepackage{makecell}

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{bm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\usepackage{graphicx}
\graphicspath{ {./img/} }

\renewcommand{\labelenumii}{\theenumii}
\renewcommand{\theenumii}{\theenumi.\arabic{enumii}.}

\begin{document}
%
\title{Application of PID controller and CNN to control Duckiebot robot}
% \title{Application of PID and CNN controllers to control the Duckiebot robot}

\author{
\IEEEauthorblockN{1\textsuperscript{st} Marek Długosz}
\IEEEauthorblockA{\textit{AGH University of Science and Technology}\\
Cracow, Poland \\
mdlugosz@agh.edu.pl \\
0000-0001-6827-9149}\\[0.4cm]
\IEEEauthorblockN{3\textsuperscript{rd} Marcin Szelest}
\IEEEauthorblockA{\textit{IEEE Senior Member}}
\IEEEauthorblockA{\textit{AGH University of Science and Technology} \\
Cracow, Poland \\
mszelest@agh.edu.pl\\
0000-0002-0522-1270}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Paweł Skruch}
\IEEEauthorblockA{\textit{IEEE Senior Member}}
\IEEEauthorblockA{\textit{AGH University of Science and Technology} \\
Cracow, Poland \\
pawel.skruch@agh.edu.pl\\
0000-0002-8290-8375}\\
\IEEEauthorblockN{4\textsuperscript{th} Artur Morys-Magiera}
\IEEEauthorblockA{\textit{AGH University of Science and Technology} \\
Cracow, Poland \\
amorys@student.agh.edu.pl\\
0000-0002-2137-8841}
}

\maketitle              % typeset the title of the contribution

\begin{abstract}

\end{abstract}

% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\begin{IEEEkeywords}
autonomouse robot, PID controller, convolutional neural network, line follower
\end{IEEEkeywords}
%
\section{Introduction}
Recent years have seen a kind of renaissance in the use of artificial neural networks. Various types of network architectures have been created and are generally available, their training methods have been improved. All this has led to the fact that artificial neural networks are increasingly used, with success, in automatic control systems \cite{algor2030973, 786109}. The use of neural networks in automation systems is not their typical use because very often control can be implemented using well-known and described classical control algorithms (PID, LQR, MPC, etc.). This provides an opportunity to compare the performance of two different technologies, to find their advantages and disadvantages. In the process of educating students, the design and implementation by them of a control system using classical control algorithms and artificial neural networks is very interesting from both a scientific and didactic point of view.

An important element of student education that increases the attractiveness of the classes themselves, providing the opportunity to solve practical problems, is the use of real robots. One such project providing a complete robotics platform (hardware, software, simulator, teaching materials) is the Duckietown project.
The Duckietown project was created in 2016 at the University of Massachusetts Institute of Technology (MIT) as part of a class for students \cite{paull2017duckietown}. It is an open project that aims to promote education and research related to robot autonomy, especially in the context of autonomous cars. As part of the project, Duckiebot robots, mock-ups of the Duckietown cities the robots navigate through, and additional infrastructure, such as the Watchtower, to support the navigation process, have been created. 
The project also includes the development of simulation software called Duckiebot-GYM, which allows simulation of the robots in a virtual environment \cite{gym_duckietown}. The second generation of Duckiebot robots is now available, using the Jetson Nano computer for control. Using the Jetson Nano platform opens the door to using advanced neural networks with dedicated GPU support to control the Duckiebot robot.

The article is organized as follows, section \ref{sec:problem-definition} defines the problem to be solved, section \ref{sec:student-activity} describes the student activities carried out in solving the problem. Section \ref{sec:robot-duckiebot} contains a brief design of Duckietown and the construction and control of the Duckiebot robot, sections \ref{sec:pid-controller} and \ref{sec:nn-controller} describe how to solve the problem using a PID controller and a controller using artificial neural networks. The \ref{sec:practical-info} section gives some practical tips to help with the task, and the final \ref{sec:conclusion} section provides conclusions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Problem definition}\label{sec:problem-definition}
As part of the didactic process, students have the goal of designing an automatic control system for a differentially driven robot that performs the functionality of moving along a designated route. The route of the robot is marked with a line of color (e.g. red, yellow, etc.) different from the ground (usually black) on which the robot moves. The shape of the route is a closed contour. The control system should generate a control signal to minimize the total deviation error of the robot from the preset route.  This type of task, is a typical sentence implemented in various types of autonomous robots and is often referred to as \emph{line follower}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Student activities}\label{sec:student-activity}
Within the problem defined in the \ref{sec:problem-definition} section, students will have to demonstrate adequate theoretical preparation and practical skills such as: 
knowledge of the principle of operation of the PID controller, knowledge of the principle of operation of artificial neural networks, knowledge of operations related to image transformations, ability to implement the discrete version of the PID controller, ability to implement selected structures of neural networks, ability to implement operations of image transformations, knowledge of issues related to the selection of parameters of the PID controller, knowledge of issues related to learning neural networks and preparing data for their learning, evaluation of the correctness of the learning process of the network. A seemingly simple algorithm for controlling an uncomplicated robot realizing the functionality of \emph{line follower} generates a number of tasks the performance of which requires adequate theoretical as well as practical preparation of students. It should also be noted that the problems that students have to solve during the project are typical problems that occur in the programming of other types of autonomous robots.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Duckiebot robot}\label{sec:robot-duckiebot}
The Duckiebot robot is a small, two-wheeled robot which is capable of moving on flat surfaces. Two modules containing an electric motor, a gearbox and an encoder were used to drive the robot. The Duckiebot robot is equipped with the following sensors: an RGB camera, a distance sensor and an inertial sensor IMU \cite{gupta2022low}. A JetsonNANO computer is used to control the robot. In addition, the Duckiebot robot is equipped with four RGB LEDs which can be used, for example, to signal the state the robot is currently in. The Duckiebot robot is designed for autonomous control and can move without human intervention, based on camera data and appropriate control algorithms. The figure \ref{fig:duckiebot-3d} shows the Duckiebot robot in three projections.

\begin{figure}[h]
    \centering
    \includegraphics[width=.9\columnwidth]{duckiebot-blue-3d}
    \caption{Duckiebot robot (\url{http://duckietown.org}).}
    \label{fig:duckiebot-3d}
\end{figure}

A consequence of using two independent electric motors in the design of the Duckiebot robot is the way in which the direction of the robot's movement is controlled.
The Duckiebot robot makes turns by changing the speeds of the electric motors depending on which direction it wants to turn and how large the turn radius should be.
Two signals are used to control the robot: $v$-value of the linear speed and $omega$-value of the rotational speed along the robot's vertical Z axis, see Fig. \ref{fig:duckiebot-kinematics}.

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\columnwidth]{dt-kinematic}
    \caption{Duckiebot robot kinematic system.}
    \label{fig:duckiebot-kinematics}
\end{figure}

Given the values of $(v, \omega)$ (determined, for example, by the controller), we can calculate the rotational speeds of the robot's wheels from the formulas \eqref{eq:omega-L} and \eqref{eq:omega-R}:

\begin{eqnarray}
	\omega_L & = & \cfrac{v-L\omega}{R} \label{eq:omega-L} \\
	\omega_R & = & \cfrac{v+L\omega}{R} \label{eq:omega-R}
\end{eqnarray}
\\
This type of Duckiebot robot control system is very commonly used in robotics and is called \emph{differential drive} \cite{siegwart2011introduction}. The advantage of such a kinematic system for the motion of the Duckiebot robot is the simplicity of the design (no torsion axis mechanisms of the wheels) while the disadvantage is the need to constantly control the rotational speed of the robot's wheels ($\omega_R$, $\omega_L$) so that it moves along the specified route.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Classic implementation with a PID controller}\label{sec:pid-controller}
The classic Duckiebot robot control system uses a PID controller. The task of the control system is to generate such a control signal $u_i$ to minimize the value of the error signal $e_i$, where $i$-consecutive time moments. Figure \ref{fig:pid-pipeline} shows a block diagram of the Duckiebot robot control system using the PID controller and the robot camera as a sensor.

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\columnwidth]{PipeLinePID}
    \caption{Schematic of Duckiebot robot control system using PID controller.}
    \label{fig:pid-pipeline}
\end{figure}

The error signal $e_i$ is defined as the difference between the center of the vertical axis of the camera image and the point in the analyzed image area containing the line along which the Duckiebot robot should move, see Fig. \ref{fig:error-definition}.

\begin{figure}[h]
    \centering
    \includegraphics[width=.95\columnwidth]{ErrorDefinition.png}
    \caption{Error determination for the control system implementing the \emph{line follower} functionality for the Duckiebot robot.}
    \label{fig:error-definition}
\end{figure}

The control system of the Duckiebot robot is a discrete system (implemented on JetsoNANO computer), so the discrete version of the PID controller has to be used. Taking into account the drive type of the Duckiebot robot \emph{differential drive robot}, the control signal $u_i$ generated by the PID controller is equivalent to the rotational speed $u_i$ along the Z axis of the robot.

\subsection{PID controller}
One of the best known and most widely used controllers in automatic control systems is the PID controller. This is primarily due to its versatility, ease of implementation in both analog and digital form, and the relatively simple and intuitive way of tuning the PID controller parameters. The discrete version of the PID controller that was used in the Duckietown robot control system is given by the equation \eqref{eq:pid}:

\begin{equation}
u_i = K_p e_i + K_i \sum_{i=0}^k{e_i \Delta t} + K_d \cfrac{e_i-e_{i-1}}{\Delta t}
\label{eq:pid}
\end{equation}
%https://github.com/duckietown/mooc-exercises/blob/c476a67fea9836beab99eed23cfc1a83c77af7bb/modcon/solution/05-PID-Control/PID_controller.ipynb
where: $K_p$ amplification of the proportional part, $K_i$ amplification of the integrating part, $K_d$ amplification of the differentiating part, $e_i$ error at the $i$th instant of time, $\Delta t$ time step \cite{aastrom2021feedback}.

\subsection{Error measurement}
The value of the error signal $e_i$ is determined from the images transmitted by the Duckiebot robot camera. In order to determine the value of the $e_i$ signal, it is necessary to perform several operations on the transmitted images (see Figure \ref{fig:pid-pipeline}). 
The first operation is to convert the image from the RGB color space to the HSV space (\emph{Hue}, \emph{Saturation}, \emph{Value}). This conversion of the color space of images is mainly aimed at better color discrimination and recognition. 
Another operation is to find a line of a certain color in the image along which the Duckiebot robot should move. Such an operation is to leave only the line marking the robot's route in the image and remove all other elements (e.g. background) from the image. 
Next operation is to reduce the area being analyzed by performing a cropping operation. This operation aims to find the exact center of the line along which the Duckiebot robot is to move. 
The center of the line is found as the coordinates of the center of gravity obtained, as a result of the previous transformations, of the line that determines the route of the Duckiebot robot. Figure \ref{fig:img-transformation-summary} shows the result of the specific transformations performed on the robot's camera image to determine the value of the position error $e_i$.

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=.95\columnwidth]{img-transformation.png}
    \caption{Operations performed on robot camera images to determine position error  $e_i$.}
    \label{fig:img-transformation-summary}
\end{figure}

Finally, the position error value $e_i$ is normalized to the interval $e_i \in [-1;1]$. This is to make independent of the size (width, height) of the images transmitted from the camera.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Implementation using neural networks}\label{sec:nn-controller}
The second method of generating a control signal $u_i$ uses a convolutional neural network (\emph{Convolutional Neural Network} CNN). The block diagram of the control system using neural networks is shown in Fig. \ref{fig:cnn-pipeline}.

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\columnwidth]{NNPipeline3}
    \caption{Schematic of Duckiebot robot control system using convolutional neural networks.}
    \label{fig:cnn-pipeline}
\end{figure}

Convolutional neural networks are very effective especially in the tasks of pattern recognition and image processing. CNNs repeatedly use the convolution operation whose use is aimed at extracting characteristic features (edges, textures, shapes) from input images on the basis of which a classical neural network can carry out an inference of "what is in the image." Convolutional neural networks are used in various fields, such as image recognition, text analysis, speech processing, facial recognition, cars, diagnosing diseases from medical images, analyzing satellite images, etc. \cite{li2021survey}, \cite{rawat2017deep}, \cite{almasi2020robust}.

When a CNN is used to control a Duckiebot robot, such a network receives images from the camera as input (after performing several transformations) based on which it generates a $u_i$ control signal which is interpreted as the $\omega_i$ speed along the robot's Z-axis, see Fig. \ref{fig:cnn-pipeline}.

\subsection{Convolutional Neural Network}
The proposed structure of the convolutional neural network is shown in Fig. \ref{fig:cnn-structure} and it should be noted that this is one of the possible network structures that can be used.

\begin{figure}[h]
    \centering
    \includegraphics[width=.5\columnwidth]{cnn2}
    \caption{Structure of the convolutional neural network used to control the Duckiebot robot.}
    \label{fig:cnn-structure}
\end{figure}

The network consists of the first convolutional layer, the next layer is the \emph{MaxPooling} layer. Its purpose is to reduce the size of the input data. The next layers of the network are the convolution layer and the \emph{MaxPooling} layer - this arrangement of layers is repeated twice. Each time, the convolution layers extract further features of the input image relevant to the control of the Duckiebot robot, while the \emph{MaxPooling} layers reduce the size of the data. The function of the next layer \emph{Dropout} is randomly turns off individual neurons of the network, so as to prevent the phenomenon of \emph{overfitting} and support the generalization ability of the neural network. Before the data is sent to the input of the actual neural network, it must be flattened, which is the function of the \emph{Flatten} layer. It transforms the multidimensional data into a one-dimensional input vector to the next layer \emph{Dense}. The \emph{Dense} layer is responsible for performing classification, regression or other machine learning tasks. The output of the CNN is the value of the signal $\omega$ in the interval $[-8: ; -8]$ therefore the activation function is from the output of the \emph{Dense} layer is the \emph{tanh} function.

\subsection{Data preparation}
Each neural network is only as good as the good data it was trained with. Hence, a very important step in working with neural networks is to prepare a good set of learning data. In the case of the Duckiebot robot and the task it is supposed to perform (\emph{line follower}), the learning data for the network consists of pairs: the image from the camera and the corresponding correct $\omega$ value, see Fig. \ref{fig:cnn-pipeline}. The data necessary to train the network can be obtained by controlling the Duckiebot robot manually (e.g., with a gamepad) or by using a previously developed control system with a PID controller by adding a script that writes the data into log file.

In addition to the appropriate amount of logged data for learning the neural network, it is important that the data is sufficiently diverse - that is, that it covers the largest possible number of cases, situations, events in which the Duckiebot robot may find itself. Figure \ref{fig:cnn-data-hist1} shows a histogram of an example dataset that has been logged.

\begin{figure}[h]
	\begin{subfigure}{.475\columnwidth}
    	\centering
    	\includegraphics[width=1\columnwidth]{h1}
    	\caption{}
    	\label{fig:cnn-data-hist1}
    \end{subfigure}
    \begin{subfigure}{.475\columnwidth}
    	\centering
    	\includegraphics[width=1\columnwidth]{h2}
    	\caption{}
    	\label{fig:cnn-data-hist2}
    \end{subfigure}

    \caption{Histogram of logged data before analysis (a) and after analysis (b).}
    \label{fig:cnn-data-hist}
\end{figure}

It can be clearly seen that the dominant value in this dataset for the $\omega$ variable is close to $0$, and if you use such data to train a CNN that is supposed to control the Duckiebot robot, it is likely that the robot would drive straight very well, while it would have problems with turns. In such a case, after analyzing the data, one should remove from the learning set data whose values significantly dominate over other values obtaining a more representative learning data set, see Fig. \ref{fig:cnn-data-hist2}.

Similar to the control of the Duckiebot robot using the PID controller, it is necessary to adequately prepare the camera images when controlling the Duckiebot robot using neural networks. The various stages of processing Duckiebot robot camera images are shown in Figure \ref{fig:cnn-pipeline} (yellow-colored blocks) and are: cropping the image to the area of interest, conversion to YUV color space (Y-luminance, UV-color, chrominance), filtering using the Gauus filter, resizing the image and normalizing the values of individual image channels to the interval $[0;1]$. The result of the various operations can be seen in Figure \ref{fig:cnn-image-prepare}.

\begin{figure}[h]
    \centering
    \includegraphics[width=.95\columnwidth]{nn_prepare_image}
    \caption{Operations performed on images from the camera of the Duckiebot robot controlled before the application of the CNN controller.}
    \label{fig:cnn-image-prepare}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Practical advice}\label{sec:practical-info}
The Duckietown robot software was implemented using the Robot Opertion System framework (\emph{ROS}) \cite{quigley2015programming}. It is currently one of the most popular frameworks for programming robots of various kinds. The consequence of choosing the ROS framework is the choice of programming language, and in the case of Duckiebot robots it is Python. The Duckietown project also defines how to develop custom software for robots and enforces the use of containerization technology. This is a very good practice in accordance with the software development trends of the time. A detailed description together with examples of creating software for Duckie robots is very well described in the documentation of the project \cite{Daniele:ff}. 
The use of the ROS framework also makes it possible to efficiently create and test software without having to constantly upload new versions to the Duckiebot robot. By configuring the host computer appropriately so that it connects to the ROS server running on the selected Duckiebot robot, you can implement and test your own software conveniently and efficiently, see Fig. \ref{fig:env-develop}.

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\columnwidth]{controll_schema2}
    \caption{Setting up a development environment for the Duckiebot robot.}
    \label{fig:env-develop}
\end{figure}

The proposed controllers should be implemented as ROS nodes. The nodes should read images from the robot's camera (sent as ROS messages) and generate a $\omega$ control signal as output. To simplify the control of the Duckiebot robot, it can be assumed that it moves at a constant linear speed $v=const$.

Convolutional neural networks can be created, trained and tested using the \emph{keras} and \emph{tensorflow} modules of the Python language. The advantage of the \emph{keras} and \emph{tensorflow} modules is also very good documentation with a very large number of examples.

It is also necessary to pay attention to the correct preparation of data for learning neural networks, each camera image must correspond to a single, correct value of the $\omega$ signal.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusions}\label{sec:conclusion}
The PID controller is used successfully in many control systems and its implementation is relatively simple. There are also a number of methods and algorithms for adjusting controller parameters for this type of controller. PID controllers are also not free of disadvantages. One of them is the requirement to know, even roughly, the model of the process we want to control. Thus, there is a need to identify both the structure of the process model and its parameters. Identification tasks are not simple tasks, requiring a great deal of knowledge about the nature of the process itself. There are also methods for identifying process models based on the results of practical experiments, but in this case it may not always be possible to conduct such an experiment. When using a PID controller, one should also be aware that it was developed for processes whose operation can be described by linear models. Unfortunately, the behavior of the vast majority of dynamic systems is described by nonlinear models. The consequence of this is that in such a case the PID controller works using linear approximations of nonlinear systems - which can cause various errors, inaccuracies, etc.

Unlike the classic PID controller, controllers using artificial neural networks do not need to know the mathematical model of the process they control and its parameters. The ability to design different neural network architectures, such as convolutional, recurrent or deep neural networks, makes it possible to adapt the neural regulator to the specific process it is supposed to control. On the other hand, the multiplicity of neural network architectures, their design means that we can never be sure whether a given neural network structure is optimal.
The selection of parameters of neural controllers is done automatically by using appropriate network learning algorithms.  The key element influencing the correctness of neural regulator operation is the data that the learned neural network is. The disadvantage of regulators using neural networks is the inability to demonstrate the stability of operation of the systems they control. In the case of the PID regulator, despite the use of approximate models of the process, it is very often possible to prove that a closed control system will operate stably in any or a certain range of values of variables. Unfortunately, such an analysis cannot be carried out in the case of neural regulators.

In summary, the implementation of two different controllers to perform the same task provides an opportunity to learn the advantages and disadvantages of each. Students, in the course of solving the problem posed, must demonstrate theoretical knowledge of the issues and, which is particularly important for engineers, acquire the ability to practically apply theoretical knowledge.

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}

### 2023-09-13 article-en.tex
English corrections

@@
 Recent years have seen a kind of renaissance in the use of artificial neural networks.
-Various types of network architectures have been created and are generally available, their training methods have been improved.
-All this has led to the fact that artificial neural networks are increasingly used, with success, in automatic control systems \cite{algor2030973, 786109}.
+Various types of network architectures have been created and are generally available, and their training methods have been improved.
+All this has led to the fact that artificial neural networks are being increasingly used, with success, in automatic control systems \cite{algor2030973, 786109}.
 The use of neural networks in automation systems is not their typical use because very often control can be implemented using well-known and described classical control algorithms (PID, LQR, MPC, etc.).
-This provides an opportunity to compare the performance of two different technologies, to find their advantages and disadvantages.
+This provides an opportunity to compare the performance of two different technologies and find their advantages and disadvantages.
 In the process of educating students, the design and implementation by them of a control system using classical control algorithms and artificial neural networks is very interesting from both a scientific and didactic point of view.
@@
 An important element of student education that increases the attractiveness of the classes themselves, providing the opportunity to solve practical problems, is the use of real robots.
-One such project providing a complete robotics platform (hardware, software, simulator, teaching materials) is the Duckietown project.
+One such project that provides a complete robotics platform (hardware, software, simulator, and teaching materials) is the Duckietown project.
 The Duckietown project was created in 2016 at the University of Massachusetts Institute of Technology (MIT) as part of a class for students \cite{paull2017duckietown}.
@@
 As part of the project, Duckiebot robots, mock-ups of the Duckietown cities the robots navigate through, and additional infrastructure, such as the Watchtower, to support the navigation process, have been created.
-The project also includes the development of simulation software called Duckiebot-GYM, which allows simulation of the robots in a virtual environment \cite{gym_duckietown}.
+The project also includes the development of simulation software called Duckiebot-GYM, which allows the simulation of robots in a virtual environment \cite{gym_duckietown}.
 The second generation of Duckiebot robots is now available, using the Jetson Nano computer for control.
@@
 
-The article is organized as follows, section \ref{sec:problem-definition} defines the problem to be solved, section \ref{sec:student-activity} describes the student activities carried out in solving the problem.
+The article is organized as follows, Section \ref{sec:problem-definition} defines the problem to be solved, Section \ref{sec:student-activity} describes the student activities carried out to solve the problem.
 Section \ref{sec:robot-duckiebot} contains a brief design of Duckietown and the construction and control of the Duckiebot robot, sections \ref{sec:pid-controller} and \ref{sec:nn-controller} describe how to solve the problem using a PID controller and a controller using artificial neural networks.
@@
 \section{Problem definition}\label{sec:problem-definition}
-As part of the didactic process, students have the goal of designing an automatic control system for a differentially driven robot that performs the functionality of moving along a designated route.
+As part of the didactic process, the students have the goal of designing an automatic control system for a differentially driven robot that performs the functionality of moving along a designated route.
 The route of the robot is marked with a line of color (e.g.
@@
 The control system should generate a control signal to minimize the total deviation error of the robot from the preset route.
-This type of task, is a typical sentence implemented in various types of autonomous robots and is often referred to as \emph{line follower}.
+This type of task is a typical sentence implemented in various types of autonomous robots and is often referred to as \emph{line follower}.
 
@@
 \section{Student activities}\label{sec:student-activity}
-Within the problem defined in the \ref{sec:problem-definition} section, students will have to demonstrate adequate theoretical preparation and practical skills such as: 
-knowledge of the principle of operation of the PID controller, knowledge of the principle of operation of artificial neural networks, knowledge of operations related to image transformations, ability to implement the discrete version of the PID controller, ability to implement selected structures of neural networks, ability to implement operations of image transformations, knowledge of issues related to the selection of parameters of the PID controller, knowledge of issues related to learning neural networks and preparing data for their learning, evaluation of the correctness of the learning process of the network.
-A seemingly simple algorithm for controlling an uncomplicated robot realizing the functionality of \emph{line follower} generates a number of tasks the performance of which requires adequate theoretical as well as practical preparation of students.
+Within the problem defined in the \ref{sec:problem-definition} section, students will have to demonstrate adequate theoretical preparation and practical skills such as: knowledge of the principle of operation of the PID controller, knowledge of the principle of operation of artificial neural networks, knowledge of operations related to image transformations, ability to implement the discrete version of the PID controller, ability to implement selected structures of neural networks, ability to implement operations of image transformations, knowledge of issues related to the selection of parameters of the PID controller, knowledge of issues related to learning neural networks and preparing data for their learning, evaluation of the correctness of the learning process of the network.
+A seemingly simple algorithm to control an uncomplicated robot that realizes the functionality of \emph{line follower} generates a number of tasks the performance of which requires adequate theoretical and practical preparation of students.
 It should also be noted that the problems that students have to solve during the project are typical problems that occur in the programming of other types of autonomous robots.
@@
 \section{Duckiebot robot}\label{sec:robot-duckiebot}
-The Duckiebot robot is a small, two-wheeled robot which is capable of moving on flat surfaces.
-Two modules containing an electric motor, a gearbox and an encoder were used to drive the robot.
-The Duckiebot robot is equipped with the following sensors: an RGB camera, a distance sensor and an inertial sensor IMU \cite{gupta2022low}.
+The Duckiebot robot is a small two-wheeled robot that is capable of moving on flat surfaces.
+Two modules containing an electric motor, a gearbox, and an encoder were used to drive the robot.
+The Duckiebot robot is equipped with the following sensors: an RGB camera, a distance sensor, and an IMU inertial sensor \cite{gupta2022low}.
 A JetsonNANO computer is used to control the robot.
@@
 This type of Duckiebot robot control system is very commonly used in robotics and is called \emph{differential drive} \cite{siegwart2011introduction}.
-The advantage of such a kinematic system for the motion of the Duckiebot robot is the simplicity of the design (no torsion axis mechanisms of the wheels) while the disadvantage is the need to constantly control the rotational speed of the robot's wheels ($\omega_R$, $\omega_L$) so that it moves along the specified route.
+The advantage of such a kinematic system for the motion of the Duckiebot robot is the simplicity of the design (no torsion axis mechanisms of the wheels), while the disadvantage is the need to constantly control the rotational speed of the robot's wheels ($\omega_R$, $\omega_L$) so that it moves along the specified route.
 
@@
 The task of the control system is to generate such a control signal $u_i$ to minimize the value of the error signal $e_i$, where $i$-consecutive time moments.
-Figure \ref{fig:pid-pipeline} shows a block diagram of the Duckiebot robot control system using the PID controller and the robot camera as a sensor.
+Figure \ref{fig:pid-pipeline} shows a block diagram of the Duckiebot robot control system that uses the PID controller and the robot camera as a sensor.
 
@@
 
-The error signal $e_i$ is defined as the difference between the center of the vertical axis of the camera image and the point in the analyzed image area containing the line along which the Duckiebot robot should move, see Fig.
+The error signal $e_i$ is defined as the difference between the center of the vertical axis of the camera image and the point in the analyzed image area containing the line along which the Duckiebot robot should move; see Fig.
 \ref{fig:error-definition}.
@@
 The value of the error signal $e_i$ is determined from the images transmitted by the Duckiebot robot camera.
-In order to determine the value of the $e_i$ signal, it is necessary to perform several operations on the transmitted images (see Figure \ref{fig:pid-pipeline}).
+To determine the value of the $e_i$ signal, it is necessary to perform several operations on the transmitted images (see Figure \ref{fig:pid-pipeline}).
 The first operation is to convert the image from the RGB color space to the HSV space (\emph{Hue}, \emph{Saturation}, \emph{Value}).
@@
 background) from the image.
-Next operation is to reduce the area being analyzed by performing a cropping operation. This operation aims to find the exact center of the line along which the Duckiebot robot is to move. 
+The next operation is to reduce the area being analyzed by performing a cropping operation. This operation aims to find the exact center of the line along which the Duckiebot robot will move. 
 The center of the line is found as the coordinates of the center of gravity obtained, as a result of the previous transformations, of the line that determines the route of the Duckiebot robot.
@@
 
-Finally, the position error value $e_i$ is normalized to the interval $e_i \in [-1;1]$. This is to make independent of the size (width, height) of the images transmitted from the camera.
+Finally, the position error value $e_i$ is normalized to the interval $e_i \in [-1;1]$. This is to be made independent of the size (width, height) of the images transmitted from the camera.
 
@@
 \section{Implementation using neural networks}\label{sec:nn-controller}
-The second method of generating a control signal $u_i$ uses a convolutional neural network (\emph{Convolutional Neural Network} CNN).
+The second method of generating a control signal $u_i$ uses a convolutional neural network (\emph{ convolutional neural network} CNN).
 The block diagram of the control system using neural networks is shown in Fig.
@@
 
-Convolutional neural networks are very effective especially in the tasks of pattern recognition and image processing.
+Convolutional neural networks are very effective, especially in the tasks of pattern recognition and image processing.
 CNNs repeatedly use the convolution operation whose use is aimed at extracting characteristic features (edges, textures, shapes) from input images on the basis of which a classical neural network can carry out an inference of "what is in the image." Convolutional neural networks are used in various fields, such as image recognition, text analysis, speech processing, facial recognition, cars, diagnosing diseases from medical images, analyzing satellite images, etc.
@@
 
-When a CNN is used to control a Duckiebot robot, such a network receives images from the camera as input (after performing several transformations) based on which it generates a $u_i$ control signal which is interpreted as the $\omega_i$ speed along the robot's Z-axis, see Fig.
+When a CNN is used to control a Duckiebot robot, such a network receives images from the camera as input (after performing several transformations) based on which it generates a $u_i$ control signal which is interpreted as the $\omega_i$ speed along the robot's Z-axis; see Fig.
 \ref{fig:cnn-pipeline}.
@@
 Its purpose is to reduce the size of the input data.
-The next layers of the network are the convolution layer and the \emph{MaxPooling} layer - this arrangement of layers is repeated twice.
+The next layers of the network are the convolution layer and the \emph{MaxPooling} layer, and this arrangement of layers is repeated twice.
 Each time, the convolution layers extract further features of the input image relevant to the control of the Duckiebot robot, while the \emph{MaxPooling} layers reduce the size of the data.
-The function of the next layer \emph{Dropout} is randomly turns off individual neurons of the network, so as to prevent the phenomenon of \emph{overfitting} and support the generalization ability of the neural network.
-Before the data is sent to the input of the actual neural network, it must be flattened, which is the function of the \emph{Flatten} layer.
-It transforms the multidimensional data into a one-dimensional input vector to the next layer \emph{Dense}.
-The \emph{Dense} layer is responsible for performing classification, regression or other machine learning tasks.
-The output of the CNN is the value of the signal $\omega$ in the interval $[-8: ; -8]$ therefore the activation function is from the output of the \emph{Dense} layer is the \emph{tanh} function.
+The function of the next layer \emph{Dropout} is randomly turns off individual neurons of the network, to prevent the phenomenon of \emph{overfitting} and support the generalization ability of the neural network.
+Before the data are sent to the input of the actual neural network, it must be flattened, which is the function of the \emph{Flatten} layer.
+Transform multidimensional data into a one-dimensional input vector to the next layer \emph{Dense}.
+The \emph{Dense} layer is responsible for performing classification, regression, or other machine learning tasks.
+The output of the CNN is the value of the signal $\omega$ in the interval $[-8: ; -8]$ therefore, the activation function is from the output of the \emph{Dense} layer is the \emph{tanh} function.
 
 \subsection{Data preparation}
-Each neural network is only as good as the good data it was trained with.
-Hence, a very important step in working with neural networks is to prepare a good set of learning data.
-In the case of the Duckiebot robot and the task it is supposed to perform (\emph{line follower}), the learning data for the network consists of pairs: the image from the camera and the corresponding correct $\omega$ value, see Fig.
+Each neural network is only as good as the good data with which it was trained.
+Therefore, a very important step in working with neural networks is to prepare a good set of learning data.
+In the case of the Duckiebot robot and the task it is supposed to perform (\emph{line follower}), the learning data for the network consist of pairs: the image from the camera and the corresponding correct $\omega$ value, see Fig.
 \ref{fig:cnn-pipeline}.
-The data necessary to train the network can be obtained by controlling the Duckiebot robot manually (e.g., with a gamepad) or by using a previously developed control system with a PID controller by adding a script that writes the data into log file.
+The data necessary to train the network can be obtained by controlling the Duckiebot robot manually (e.g., with a gamepad) or by using a previously developed control system with a PID controller by adding a script that writes the data into a log file.
 
-In addition to the appropriate amount of logged data for learning the neural network, it is important that the data is sufficiently diverse - that is, that it covers the largest possible number of cases, situations, events in which the Duckiebot robot may find itself.
+In addition to the appropriate amount of logged data for learning the neural network, it is important that the data is sufficiently diverse, that is, that it covers the largest possible number of cases, situations, events in which the Duckiebot robot may find itself.
 Figure \ref{fig:cnn-data-hist1} shows a histogram of an example dataset that has been logged.
@@
 
-    \caption{Histogram of logged data before analysis (a) and after analysis (b).}
+    \caption{Histogram of data recorded before analysis (a) and after analysis (b).}
     \label{fig:cnn-data-hist}
@@
 
-It can be clearly seen that the dominant value in this dataset for the $\omega$ variable is close to $0$, and if you use such data to train a CNN that is supposed to control the Duckiebot robot, it is likely that the robot would drive straight very well, while it would have problems with turns.
-In such a case, after analyzing the data, one should remove from the learning set data whose values significantly dominate over other values obtaining a more representative learning data set, see Fig.
+It can be clearly seen that the dominant value in this data set for the variable $\omega$ is close to $0$, and if you use such data to train a CNN that is supposed to control the Duckiebot robot, it is likely that the robot would drive straight very well, while it would have problems turning.
+In such a case, after analyzing the data, data whose values significantly dominate other values should be removed from the learning set, obtaining a more representative learning data set, see Fig.
 \ref{fig:cnn-data-hist2}.
 
-Similar to the control of the Duckiebot robot using the PID controller, it is necessary to adequately prepare the camera images when controlling the Duckiebot robot using neural networks.
-The various stages of processing Duckiebot robot camera images are shown in Figure \ref{fig:cnn-pipeline} (yellow-colored blocks) and are: cropping the image to the area of interest, conversion to YUV color space (Y-luminance, UV-color, chrominance), filtering using the Gauus filter, resizing the image and normalizing the values of individual image channels to the interval $[0;1]$.
+Similarly to the control of the Duckiebot robot using the PID controller, it is necessary to adequately prepare the camera images when controlling the Duckiebot robot using neural networks.
+The various stages of processing Duckiebot robot camera images are shown in Figure \ref{fig:cnn-pipeline} (yellow blocks) and are: cropping the image to the area of interest, conversion to YUV color space (Y-luminance, UV color, chrominance), filtering using the Gauus filter, resizing the image and normalizing the values of individual image channels to the interval $[0;1]$.
 The result of the various operations can be seen in Figure \ref{fig:cnn-image-prepare}.
@@
 This is a very good practice in accordance with the software development trends of the time.
-A detailed description together with examples of creating software for Duckie robots is very well described in the documentation of the project \cite{Daniele:ff}.
+A detailed description together with examples of creating software for Duckie robots is very well described in the project documentation \cite{Daniele:ff}.
 The use of the ROS framework also makes it possible to efficiently create and test software without having to constantly upload new versions to the Duckiebot robot.
@@
 The proposed controllers should be implemented as ROS nodes.
-The nodes should read images from the robot's camera (sent as ROS messages) and generate a $\omega$ control signal as output.
+Nodes should read images from the robot's camera (sent as ROS messages) and generate a $\omega$ control signal as output.
 To simplify the control of the Duckiebot robot, it can be assumed that it moves at a constant linear speed $v=const$.
 
-Convolutional neural networks can be created, trained and tested using the \emph{keras} and \emph{tensorflow} modules of the Python language.
+Convolutional neural networks can be created, trained, and tested using the \emph{keras} and \emph{tensorflow} modules of the Python language.
 The advantage of the \emph{keras} and \emph{tensorflow} modules is also very good documentation with a very large number of examples.
@@
 \section{Conclusions}\label{sec:conclusion}
-The PID controller is used successfully in many control systems and its implementation is relatively simple.
+The PID controller is used successfully in many control systems, and its implementation is relatively simple.
 There are also a number of methods and algorithms for adjusting controller parameters for this type of controller.
 PID controllers are also not free of disadvantages.
-One of them is the requirement to know, even roughly, the model of the process we want to control.
-Thus, there is a need to identify both the structure of the process model and its parameters.
+One of them is the requirement of knowing, even roughly, the model of the process we want to control.
+Thus, it is necessary to identify both the structure of the process model and its parameters.
 Identification tasks are not simple tasks, requiring a great deal of knowledge about the nature of the process itself.
@@
 When using a PID controller, one should also be aware that it was developed for processes whose operation can be described by linear models.
-Unfortunately, the behavior of the vast majority of dynamic systems is described by nonlinear models.
-The consequence of this is that in such a case the PID controller works using linear approximations of nonlinear systems - which can cause various errors, inaccuracies, etc.
+Unfortunately, the behavior of the vast majority of dynamic systems is described by non-linear models.
+The consequence of this is that in such a case the PID controller works using linear approximations of nonlinear systems, which can cause various errors, inaccuracies, etc.
 
 Unlike the classic PID controller, controllers using artificial neural networks do not need to know the mathematical model of the process they control and its parameters.
-The ability to design different neural network architectures, such as convolutional, recurrent or deep neural networks, makes it possible to adapt the neural regulator to the specific process it is supposed to control.
-On the other hand, the multiplicity of neural network architectures, their design means that we can never be sure whether a given neural network structure is optimal.
-The selection of parameters of neural controllers is done automatically by using appropriate network learning algorithms.
-The key element influencing the correctness of neural regulator operation is the data that the learned neural network is.
+The ability to design different neural network architectures, such as convolutional, recurrent, or deep neural networks, makes it possible to adapt the neural regulator to the specific process it is supposed to control.
+On the other hand, the multiplicity of neural network architectures and their design means that we can never be sure whether a given neural network structure is optimal.
+The selection of neural controller parameters is done automatically using appropriate network learning algorithms.
+The key element influencing the accuracy of neural regulator operation is the data that the learned neural network is.
 The disadvantage of regulators using neural networks is the inability to demonstrate the stability of operation of the systems they control.

### 2023-09-13 article-en.tex
Add polish and english abstract

@@
 \begin{abstract}
-
+The paper presents the design and practical implementation by students of a control system
+using a classical PID controller and a controller using artificial neural networks.
+The control object is a Duckiebot robot and the task it is to perform is to drive the robot along a designated line (\emph{line follower}).
+The purpose of the proposed activities is to familiarize students with the advantages and disadvantages of the two controllers used and for them to acquire the ability to implement control systems in practice.
+The article briefly describes how the two controllers work, how to practically implement them, and describes how to practically implement the exercise. 
 \end{abstract}

### 2023-09-13 article-pl.tex
Add polish and english abstract

@@
 \begin{abstract}
-
+W artykule przedstawiono projekt i praktyczną realizację przez studentów układu sterowania
+z wykorzystaniem klasycznego regulatora PID i regulatora wykorzystującego sztuczne sieci neurnowe.
+Obiektem sterowania jest robot Duckiebot a zadanie jakie ma realizować polega na jeździe robota wzdłuż wyznaczonej lini (\emph{linie follower}).
+Celem proponowanych aktywności jest zapoznanie studentów z zaletami  i wadami obydwu użytych regulatorów oraz nabycia przez nich umiejętności praktycznej realizacji układów sterowania.
+W artykule krótko opisano sposób działania obydwu regulatorów, sposób ich praktycznej implementacji oraz opisano w jaki sposób praktycznie należy zrealizować ćwiczenie. 
 \end{abstract}

### 2023-09-13 article-en.tex
English abstract correction

--- 2023-09-13 article-en.tex
@@
 \begin{abstract}
-
+The paper presents the design and practical implementation by students of a control system
+using a classical PID controller and a controller using artificial neural networks.
+The control object is a Duckiebot robot, and the task it is to perform is to drive the robot along a designated line (\emph{line follower}).
+The purpose of the proposed activities is to familiarize students with the advantages and disadvantages of the two controllers used and to acquire the ability to implement control systems in practice.
+The article briefly describes how the two controllers work, how to practically implement them, and how to practically implement the exercise. 
 \end{abstract}

### 2023-09-20 article-en.tex
English version correction

--- 2023-09-13 article-en.tex
@@
     urlcolor=cyan,
-    pdftitle={Overleaf Example},
+    pdftitle={Application of PID controller and CNN to control Duckiebot robot},
     pdfpagemode=FullScreen,
@@
 \begin{abstract}
-
+The paper presents the design and practical implementation by students of a control system
+using a classic PID controller and a controller using artificial neural networks.
+The control object is a Duckiebot robot, and the task it is to perform is to drive the robot along a designated line (\emph{line follower}).
+The purpose of the proposed activities is to familiarize students with the advantages and disadvantages of the two controllers used and for them to acquire the ability to implement control systems in practice.
+The article briefly describes how the two controllers work, how to practically implement them, and how to practically implement the exercise.
 \end{abstract}
@@
 \section{Introduction}
-Recent years have seen a kind of renaissance in the use of artificial neural networks.
-Various types of network architectures have been created and are generally available, and their training methods have been improved.
-All this has led to the fact that artificial neural networks are being increasingly used, with success, in automatic control systems \cite{algor2030973, 786109}.
-The use of neural networks in automation systems is not their typical use because very often control can be implemented using well-known and described classical control algorithms (PID, LQR, MPC, etc.).
+Recent years have seen a renaissance of sorts in the use of artificial neural networks.
+Various types of network architectures have been created and made generally available, as well as their training methods have been improved.
+All this has led to the application of artificial neural networks being on a rise, particularly successfully, in automatic control systems \cite{algor2030973, 786109}.
+The use of neural networks in automation systems is not their typical application because very often control can be implemented using well-known and described classical control algorithms (PID, LQR, MPC, etc.).
 This provides an opportunity to compare the performance of two different technologies and find their advantages and disadvantages.
@@
 An important element of student education that increases the attractiveness of the classes themselves, providing the opportunity to solve practical problems, is the use of real robots.
-One such project that provides a complete robotics platform (hardware, software, simulator, and teaching materials) is the Duckietown project.
-The Duckietown project was created in 2016 at the University of Massachusetts Institute of Technology (MIT) as part of a class for students \cite{paull2017duckietown}.
+One project that provides a complete robotics platform (hardware, software, simulator, and teaching materials) is the Duckietown project.
+The Duckietown project was created in 2016 at the Massachusetts Institute of Technology (MIT) as part of a class for students \cite{paull2017duckietown}.
 It is an open project that aims to promote education and research related to robot autonomy, especially in the context of autonomous cars.
-As part of the project, Duckiebot robots, mock-ups of the Duckietown cities the robots navigate through, and additional infrastructure, such as the Watchtower, to support the navigation process, have been created.
+As part of the project, the authors designed entities like Duckiebot robots, mock-ups of the Duckietown cities the robots need to navigate through, and additional infrastructure, such as the Watchtowers, to support the navigation process.
 The project also includes the development of simulation software called Duckiebot-GYM, which allows the simulation of robots in a virtual environment \cite{gym_duckietown}.
-The second generation of Duckiebot robots is now available, using the Jetson Nano computer for control.
+The most recent, second generation of Duckiebot robots, is using the Jetson Nano single-board computer for control.
 Using the Jetson Nano platform opens the door to using advanced neural networks with dedicated GPU support to control the Duckiebot robot.
 
-The article is organized as follows, Section \ref{sec:problem-definition} defines the problem to be solved, Section \ref{sec:student-activity} describes the student activities carried out to solve the problem.
-Section \ref{sec:robot-duckiebot} contains a brief design of Duckietown and the construction and control of the Duckiebot robot, sections \ref{sec:pid-controller} and \ref{sec:nn-controller} describe how to solve the problem using a PID controller and a controller using artificial neural networks.
-The \ref{sec:practical-info} section gives some practical tips to help with the task, and the final \ref{sec:conclusion} section provides conclusions.
+The article is organized as follows: Section \ref{sec:problem-definition} defines the problem to be solved, Section \ref{sec:student-activity} describes the student activities that need to be carried out to solve the problem.
+Section \ref{sec:robot-duckiebot} contains a brief design of Duckietown, the structure and control of the Duckiebot robot, sections \ref{sec:pid-controller} and \ref{sec:nn-controller} describe how to solve the problem using a PID controller and a controller using artificial neural networks.
+The \ref{sec:practical-info} section gives some practical tips to help with the task, and finally the \ref{sec:conclusion} section provides conclusions.
 
@@
 \section{Problem definition}\label{sec:problem-definition}
-As part of the didactic process, the students have the goal of designing an automatic control system for a differentially driven robot that performs the functionality of moving along a designated route.
-The route of the robot is marked with a line of color (e.g.
-red, yellow, etc.) different from the ground (usually black) on which the robot moves.
-The shape of the route is a closed contour.
-The control system should generate a control signal to minimize the total deviation error of the robot from the preset route.
-This type of task is a typical sentence implemented in various types of autonomous robots and is often referred to as \emph{line follower}.
+As part of the didactic process, students have the goal of designing an automatic control system for a differentially driven robot that performs the functionality of moving along a designated route.
+The route of the robot is marked with a colored line (e.g.
+red, yellow, etc.) distinguishable from the ground (usually black), on which the robot moves.
+The shape of the trajectory is a closed contour.
+The control system should generate a control signal to minimize the total deviation error of the robot from the preset trajectory.
+This type of task is a typical scenario implemented in various types of autonomous robots and is often referred to as the \emph{line follower}.
 
@@
 \section{Student activities}\label{sec:student-activity}
-Within the problem defined in the \ref{sec:problem-definition} section, students will have to demonstrate adequate theoretical preparation and practical skills such as: knowledge of the principle of operation of the PID controller, knowledge of the principle of operation of artificial neural networks, knowledge of operations related to image transformations, ability to implement the discrete version of the PID controller, ability to implement selected structures of neural networks, ability to implement operations of image transformations, knowledge of issues related to the selection of parameters of the PID controller, knowledge of issues related to learning neural networks and preparing data for their learning, evaluation of the correctness of the learning process of the network.
-A seemingly simple algorithm to control an uncomplicated robot that realizes the functionality of \emph{line follower} generates a number of tasks the performance of which requires adequate theoretical and practical preparation of students.
-It should also be noted that the problems that students have to solve during the project are typical problems that occur in the programming of other types of autonomous robots.
+Within the problem defined in the \ref{sec:problem-definition} section, students will have to demonstrate adequate theoretical preparation and practical skills such as: knowledge of the principle of operation of the PID controller, knowledge of the principle of operation of artificial neural networks, knowledge of operations related to image transformations, ability to implement the discrete version of the PID controller, skills to implement selected structures of neural networks, ability to implement operations of image transformations, knowledge of issues related to the selection of parameters of the PID controller, knowledge of issues related to training neural networks and preparing data for their training, evaluation of the correctness of the training process of the network.
+A seemingly simple algorithm to control an uncomplicated robot that realizes the functionality of \emph{line follower} generates a number of tasks, performance of which requires adequate theoretical and practical preparation from students.
+It should also be noted that the problems that students have to solve during the project are typical problems that also occur in the programming of other types of autonomous robots.
 
@@
 The Duckiebot robot is a small two-wheeled robot that is capable of moving on flat surfaces.
-Two modules containing an electric motor, a gearbox, and an encoder were used to drive the robot.
-The Duckiebot robot is equipped with the following sensors: an RGB camera, a distance sensor, and an IMU inertial sensor \cite{gupta2022low}.
-A JetsonNANO computer is used to control the robot.
-In addition, the Duckiebot robot is equipped with four RGB LEDs which can be used, for example, to signal the state the robot is currently in.
+Two modules containing an electric motor, a gearbox and an encoder were used to drive the robot.
+The Duckiebot robot is equipped with the following sensors: an RGB camera, a distance sensor and an IMU inertial sensor \cite{gupta2022low}.
+A Jetson NANO single-board computer is used to control the robot.
+In addition, the Duckiebot robot is equipped with four RGB LEDs which can be used, for instance, to signal the state the robot is currently in.
 The Duckiebot robot is designed for autonomous control and can move without human intervention, based on camera data and appropriate control algorithms.
@@
 A consequence of using two independent electric motors in the design of the Duckiebot robot is the way in which the direction of the robot's movement is controlled.
-The Duckiebot robot makes turns by changing the speeds of the electric motors depending on which direction it wants to turn and how large the turn radius should be.
+The Duckiebot robot makes turns by changing the speeds of the electric motors depending on which direction it wants to turn and how large the turning radius should be.
 Two signals are used to control the robot: $v$-value of the linear speed and $omega$-value of the rotational speed along the robot's vertical Z axis, see Fig. \ref{fig:duckiebot-kinematics}.
@@
 The classic Duckiebot robot control system uses a PID controller.
-The task of the control system is to generate such a control signal $u_i$ to minimize the value of the error signal $e_i$, where $i$-consecutive time moments.
+The task of the control system is to generate such a control signal $u_i$ to minimize the value of the error signal $e_i$, where $i$ corresponds to consecutive time moments.
 Figure \ref{fig:pid-pipeline} shows a block diagram of the Duckiebot robot control system that uses the PID controller and the robot camera as a sensor.
@@
 
-The error signal $e_i$ is defined as the difference between the center of the vertical axis of the camera image and the point in the analyzed image area containing the line along which the Duckiebot robot should move; see Fig.
+The error signal $e_i$ is defined as the difference between the center of the horizontal axis of the camera image and the point in the analyzed image region contained within the line along which the Duckiebot robot should move; see Fig.
 \ref{fig:error-definition}.
@@
 
-The control system of the Duckiebot robot is a discrete system (implemented on JetsoNANO computer), so the discrete version of the PID controller has to be used.
+The control system of the Duckiebot robot is a discrete system (implemented on Jetson NANO computer), so the discrete version of the PID controller has to be used.
 Taking into account the drive type of the Duckiebot robot \emph{differential drive robot}, the control signal $u_i$ generated by the PID controller is equivalent to the rotational speed $u_i$ along the Z axis of the robot.
@@
 %https://github.com/duckietown/mooc-exercises/blob/c476a67fea9836beab99eed23cfc1a83c77af7bb/modcon/solution/05-PID-Control/PID_controller.ipynb
-where: $K_p$ amplification of the proportional part, $K_i$ amplification of the integrating part, $K_d$ amplification of the differentiating part, $e_i$ error at the $i$th instant of time, $\Delta t$ time step \cite{aastrom2021feedback}.
+where: $K_p$ amplification of the proportional part, $K_i$ amplification of the integrating part, $K_d$ amplification of the differentiating part, $e_i$ error at the $i$-th instant of time, $\Delta t$ time step \cite{aastrom2021feedback}.
 
@@
 The first operation is to convert the image from the RGB color space to the HSV space (\emph{Hue}, \emph{Saturation}, \emph{Value}).
-This conversion of the color space of images is mainly aimed at better color discrimination and recognition.
-Another operation is to find a line of a certain color in the image along which the Duckiebot robot should move.
+This conversion of between color spaces of images is mainly aimed at better color discrimination and recognition.
+Another operation is to find a line of certain color in the image, along which the Duckiebot robot should move.
 Such an operation is to leave only the line marking the robot's route in the image and remove all other elements (e.g.
@@
 
-Finally, the position error value $e_i$ is normalized to the interval $e_i \in [-1;1]$. This is to be made independent of the size (width, height) of the images transmitted from the camera.
+Finally, the position error value $e_i$ is normalized to the interval $e_i \in [-1;1]$, so as to make the result independent of the resolution (width, height) of the images captured from the camera.
 
@@
 \section{Implementation using neural networks}\label{sec:nn-controller}
-The second method of generating a control signal $u_i$ uses a convolutional neural network (\emph{ convolutional neural network} CNN).
+The second method of generating a control signal $u_i$ uses a convolutional neural network (\emph{convolutional neural network} CNN).
 The block diagram of the control system using neural networks is shown in Fig.
@@
 
-Convolutional neural networks are very effective, especially in the tasks of pattern recognition and image processing.
-CNNs repeatedly use the convolution operation whose use is aimed at extracting characteristic features (edges, textures, shapes) from input images on the basis of which a classical neural network can carry out an inference of "what is in the image." Convolutional neural networks are used in various fields, such as image recognition, text analysis, speech processing, facial recognition, cars, diagnosing diseases from medical images, analyzing satellite images, etc.
+Convolutional neural networks are very effective in the tasks of pattern recognition and image processing.
+CNNs repeatedly use the convolution operation, which allows for extracting characteristic features (edges, textures, shapes) from input images on the basis of which a classical neural network can judge on "what is in the image".
+Convolutional neural networks are used in various fields, such as image recognition, text analysis, speech processing, facial recognition, control of vehicles, diagnosing diseases from medical images, analyzing satellite images, etc.
 \cite{li2021survey}, \cite{rawat2017deep}, \cite{almasi2020robust}.
@@
 The next layers of the network are the convolution layer and the \emph{MaxPooling} layer, and this arrangement of layers is repeated twice.
-Each time, the convolution layers extract further features of the input image relevant to the control of the Duckiebot robot, while the \emph{MaxPooling} layers reduce the size of the data.
-The function of the next layer \emph{Dropout} is randomly turns off individual neurons of the network, to prevent the phenomenon of \emph{overfitting} and support the generalization ability of the neural network.
-Before the data are sent to the input of the actual neural network, it must be flattened, which is the function of the \emph{Flatten} layer.
-Transform multidimensional data into a one-dimensional input vector to the next layer \emph{Dense}.
-The \emph{Dense} layer is responsible for performing classification, regression, or other machine learning tasks.
-The output of the CNN is the value of the signal $\omega$ in the interval $[-8: ; -8]$ therefore, the activation function is from the output of the \emph{Dense} layer is the \emph{tanh} function.
+Each time, the convolution layers extract further features of the input image relevant to the control of the Duckiebot robot, while the \emph{MaxPooling} layers reduce the size of the feature maps.
+The function of the next layer \emph{Dropout} is randomly turning off individual neurons of the network, to prevent the phenomenon of \emph{overfitting} and to further support the generalization ability of the neural network.
+Before the data is sent to the input of the actual neural network, it must be flattened, which is the reasoning behind placing a \emph{Flatten} layer: to transform multidimensional data into a one-dimensional input vector to pass to the next, \emph{Dense} layer.
+The \emph{Dense} layer is responsible for performing abstract data representation tasks, e.g.
+may be used for classification, regression.
+The output of the CNN is the value of the signal $\omega$ in the interval $[-8 ; -8]$, therefore, the activation function from the \emph{Dense} output layer is the \emph{tanh} function.
 
 \subsection{Data preparation}
-Each neural network is only as good as the good data with which it was trained.
-Therefore, a very important step in working with neural networks is to prepare a good set of learning data.
-In the case of the Duckiebot robot and the task it is supposed to perform (\emph{line follower}), the learning data for the network consist of pairs: the image from the camera and the corresponding correct $\omega$ value, see Fig.
+A neural network is only as good as the data with which it was trained.
+Therefore, a very important step in working with neural networks is to prepare a good set of training data.
+In case of the Duckiebot robot and the task it is supposed to perform (\emph{line follower}), the training data for the network consist of pairs: the image from the camera and the corresponding correct $\omega$ value, see Fig.
 \ref{fig:cnn-pipeline}.
@@
 
-In addition to the appropriate amount of logged data for learning the neural network, it is important that the data is sufficiently diverse, that is, that it covers the largest possible number of cases, situations, events in which the Duckiebot robot may find itself.
+In addition to the appropriate amount of logged data for training the neural network, it is important that the data is sufficiently diverse, that is, that it covers the largest possible number of cases, situations, events in which the Duckiebot robot may find itself.
 Figure \ref{fig:cnn-data-hist1} shows a histogram of an example dataset that has been logged.
@@
 It can be clearly seen that the dominant value in this data set for the variable $\omega$ is close to $0$, and if you use such data to train a CNN that is supposed to control the Duckiebot robot, it is likely that the robot would drive straight very well, while it would have problems turning.
-In such a case, after analyzing the data, data whose values significantly dominate other values should be removed from the learning set, obtaining a more representative learning data set, see Fig.
+In such a case, after analyzing the data, samples with values that significantly dominate other values should be discarded from the training set, obtaining a more representative training data set, see Fig.
 \ref{fig:cnn-data-hist2}.
 
-Similarly to the control of the Duckiebot robot using the PID controller, it is necessary to adequately prepare the camera images when controlling the Duckiebot robot using neural networks.
-The various stages of processing Duckiebot robot camera images are shown in Figure \ref{fig:cnn-pipeline} (yellow blocks) and are: cropping the image to the area of interest, conversion to YUV color space (Y-luminance, UV color, chrominance), filtering using the Gauus filter, resizing the image and normalizing the values of individual image channels to the interval $[0;1]$.
+Similarly to control of the Duckiebot robot using a PID controller, it is necessary to adequately prepare the camera images when controlling the Duckiebot robot using neural networks.
+The various stages of processing Duckiebot robot camera images are shown in Figure \ref{fig:cnn-pipeline} (yellow blocks) and are: cropping the image to the area of interest, conversion to YUV color space (Y-luminance, UV color, chrominance), filtering using the Gauss filter, resizing the image and normalizing the values of individual image channels to the interval $[0;1]$.
 The result of the various operations can be seen in Figure \ref{fig:cnn-image-prepare}.
@@
 The Duckietown project also defines how to develop custom software for robots and enforces the use of containerization technology.
-This is a very good practice in accordance with the software development trends of the time.
+This is a very good practice in accordance with the current software development trends.
 A detailed description together with examples of creating software for Duckie robots is very well described in the project documentation \cite{Daniele:ff}.
@@
 
-Convolutional neural networks can be created, trained, and tested using the \emph{keras} and \emph{tensorflow} modules of the Python language.
-The advantage of the \emph{keras} and \emph{tensorflow} modules is also very good documentation with a very large number of examples.
+Convolutional neural networks can be created, trained, and tested using the \emph{keras} and \emph{tensorflow} libraries in the Python programming language.
+The advantage of the \emph{keras} and \emph{tensorflow} modules is also a very good documentation with a very large number of examples, as well as a large community of developers gathered around them.
 
-It is also necessary to pay attention to the correct preparation of data for learning neural networks, each camera image must correspond to a single, correct value of the $\omega$ signal.
+It is also necessary to pay attention to the correct preparation of data for training neural networks, each camera image must correspond to a single, correct value of the $\omega$ signal.
 
@@
 There are also a number of methods and algorithms for adjusting controller parameters for this type of controller.
-PID controllers are also not free of disadvantages.
-One of them is the requirement of knowing, even roughly, the model of the process we want to control.
+PID controllers, on the other hand, are not free of disadvantages.
+One of them is the requirement of prior knowledge of, even roughly, the model of the process one wants to control.
 Thus, it is necessary to identify both the structure of the process model and its parameters.
-Identification tasks are not simple tasks, requiring a great deal of knowledge about the nature of the process itself.
-There are also methods for identifying process models based on the results of practical experiments, but in this case it may not always be possible to conduct such an experiment.
-When using a PID controller, one should also be aware that it was developed for processes whose operation can be described by linear models.
+Identification tasks are complex tasks, requiring a great deal of knowledge about the nature of the process itself.
+There are also methods for identifying process models based on the results of practical experiments, however sometimes it may not be possible to conduct such experiments.
+When using a PID controller, one should also be aware that it was developed for processes, operation of which can be described by linear models.
 Unfortunately, the behavior of the vast majority of dynamic systems is described by non-linear models.
-The consequence of this is that in such a case the PID controller works using linear approximations of nonlinear systems, which can cause various errors, inaccuracies, etc.
+The consequence of this fact is that, in such cases, the PID controller works using linear approximations of nonlinear systems, which can lead to various errors, inaccuracies, etc.
 
@@
 On the other hand, the multiplicity of neural network architectures and their design means that we can never be sure whether a given neural network structure is optimal.
-The selection of neural controller parameters is done automatically using appropriate network learning algorithms.
-The key element influencing the accuracy of neural regulator operation is the data that the learned neural network is.
+The selection of neural controller parameters is done automatically using appropriate network training algorithms.
+The key element influencing the accuracy of neural regulator operation is the data used for training the neural network.
 The disadvantage of regulators using neural networks is the inability to demonstrate the stability of operation of the systems they control.
-In the case of the PID regulator, despite the use of approximate models of the process, it is very often possible to prove that a closed control system will operate stably in any or a certain range of values of variables.
+In case of the PID regulator, despite the use of approximate models of the process, it is very often possible to prove that a closed control system will operate stably in any or a certain range of values of variables.
 Unfortunately, such an analysis cannot be carried out in the case of neural regulators.
@@
 In summary, the implementation of two different controllers to perform the same task provides an opportunity to learn the advantages and disadvantages of each.
-Students, in the course of solving the problem posed, must demonstrate theoretical knowledge of the issues and, which is particularly important for engineers, acquire the ability to practically apply theoretical knowledge.
+Students, in the course of solving the problem presented, must demonstrate theoretical knowledge of the issues and - which is particularly important for engineers - acquire the ability to practically apply theoretical knowledge.
 
